Comparing Optimization Algorithms in the Fitting of Linear Mixed Models: Evaluating Speed and Accuracy using \pkg{lme4} in R and \pkg{lmm} in Julia

### 2015-08-03 inst/doc/Paper.Rnw
edits to methods section

>>>>

To provide consistency we have copied all the data sets used in the
timings to the \pkg{Timings} package itself. We have done all timings
on the same computer. This computer has a relatively recent Intel
processor and we used the
\href{https://software.intel.com/en-us/intel-mkl}{Intel Math Kernel
Library (MKL)} with \proglang{Julia}. We attempted to use
\href{www.revolutionanalytics.com/revolution-r-open}{Revolution R Open
(RRO)} as the \proglang{R} implementation as it can be configured with \pkg{MKL}.
However, we ran into version problems with this so we used the standard
Ubuntu version of \proglang{R} linked against OpenBLAS, which is also
multi-threaded.

Variables were renamed in the pattern:

\begin{itemize}
\item \textbf{Y} the response 
\item \textbf{A}, \textbf{B}, \dots{} categorical covariates 
\item \textbf{G},\textbf{H}, \textbf{I}, \dots{} grouping factors for random effects 
\item \textbf{U}, \textbf{V}, \dots{} (skipping \textbf{Y}) continuous covariates
\end{itemize}

The timing results are saved in \href{http://json.org}{JSON (JavaScript
Object Notation)} files in the directory accessible as

<<eval=FALSE>>=
system.file("JSON",package="Timings")
@

within \pkg{R}. The directory name will end with
\texttt{./Timings/inst/JSON/} in the package source directory, for
example the result of cloning the
\href{https://github.com/Stat990-033/Timings}{github repository}. There
is one \texttt{.json} file for each data set. Each such file contains
results on timings of one or more models.

The \pkg{Timings} package for \proglang{R} provides a \texttt{retime}
function that takes the name of one of these JSON files and, optionally,
the name of a file with the updated timings. Similarly there are some
source files for Julia retimings.

%    \begin{Verbatim}[commandchars=\\\{\}]
%{\color{incolor}In [{\color{incolor}1}]:} \PY{n}{include}\PY{p}{(}\PY{l+s}{\PYZdq{}}\PY{l+s}{../julia%%/retime.jl}\PY{l+s}{\PYZdq{}}\PY{p}{)}
%        \PY{n}{retime}\PY{p}{(}\PY{l+s}{\PYZdq{}}\PY{l+s}{../JSON/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p%}{,}\PY{l+s}{\PYZdq{}}\PY{l+s}{/tmp/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p}{)}\PY{p}{;}
%       \PY{n}{retime}\PY{p}{(}\PY{l+s}{\PYZdq{}}\PY{l+s}{../JSON/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p}%{,}\PY{l+s}{\PYZdq{}}\PY{l+s}{/tmp/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p}{)}\PY{p}{;}
%\end{Verbatim}

    \begin{Verbatim}[commandchars=\\\{\}]
dsname => "Alfalfa"
    \end{Verbatim}

    \begin{Verbatim}[commandchars=\\\{\}]

        `parse` has no method matching parse(::Int64)
    while loading In[1], in expression starting on line 2

         in retime at /home/bates/git/Timings/inst/julia/retime.jl:17

    \end{Verbatim}

    The timing was repeated so that compilation time is not included in the
results. This repetition is only needed once per session.

A careful examination of these results shows that the main differences
in the Julia timings (the R timings are merely reported, not evaluated)
are that the \texttt{LN\_BOBYQA} and \texttt{LD\_MMA} optimizers are
much faster in the second run. This is because much of the code needs to
be compiled the first time that a derivative-free optimizer and a
derivative-based optimizer are used.

The names of the optimizers used with \texttt{lmm} are those from the
\href{https://github.com/JuliaOpt/NLopt.jl}{NLopt package} for
\textbf{Julia}. Names that begin with \texttt{LD\_} are gradient-based
methods. Names that begin with \texttt{LN\_} are derivative-free
methods. There is one other derivative-free method, \texttt{LN\_PRAXIS},
available in the \texttt{NLopt} package but, for some reason, it can
hang on very simple problems like this. Frequently we omit it.

The optimizers used with \texttt{lmer} include the \texttt{Nelder\_Mead}
optimizer built into the \texttt{lme4} package, the \texttt{bobyqa}
optimizer from the
\href{http://cran.rstudio.com/web/packages/minqa/index.html}{minqa
package}, the derivative-free optimizers from the
\href{http://cran.rstudio.com/web/packages/nloptr/index.html}{nloptr
package} and several optimizers from the
\href{http://cran.rstudio.com/web/packages/optimx/index.html}{optimx
pacakge}.

The \texttt{optimx:bobyqa} optimizer is just a wrapper around
\texttt{bobyqa} (bounded optimization by quadratic approximation) from
the \texttt{minqa} package and should provide results similar to those
from the \texttt{bobyqa} optimizer. For some reason the number of
function evaluations is not reported for the version in \texttt{optimx}.

The optimizers from \texttt{nloptr} (i.e.~those whose names begin with
\texttt{NLOPT\_LN\_}) use the same underlying code as do the similarly
named optimizers in the \texttt{NLopt} package for \textbf{Julia}. The
number of iterations to convergence should be similar for the same
underlying code, although not nessarily exactly the same because the
evaluation of the objective in \textbf{R} and in \textbf{Julia} may
produce slightly different answers. Also the convergence criteria in the
\textbf{Julia} version are more strict than those in the \textbf{R}
version

Also shown are the value of the criterion (negative twice the
log-likelihood, lower is better) achieved, the elapsed time and the
number of function and gradient evaluations. The \texttt{nopt} value is
the number of parameters in the optimization problem. \texttt{mtype} is
the model type in the \textbf{Julia} code. There are special methods for
solving the penalized least squares (PLS) problem, and for evaluating
the objective and its gradient when there is only one grouping factor
for the random effects. The model type is called \texttt{PLSOne}.

The \texttt{Alfalfa} example is a particularly easy one and all of the
optimizerws converge to an objective value close to -10.81023 in less
than 0.6 seconds.

### 2015-08-03 inst/doc/Paper.tex
edits to methods section

>>>>

To provide consistency we have copied all the data sets used in the
timings to the \pkg{Timings} package itself. We have done all timings
on the same computer. This computer has a relatively recent Intel
processor and we used the
\href{https://software.intel.com/en-us/intel-mkl}{Intel Math Kernel
Library (MKL)} with \proglang{Julia}. We attempted to use
\href{www.revolutionanalytics.com/revolution-r-open}{Revolution R Open
(RRO)} as the \proglang{R} implementation as it can be configured with \pkg{MKL}.
However, we ran into version problems with this so we used the standard
Ubuntu version of R linked against OpenBLAS, which is also
multi-threaded.

Variables were renamed in the pattern:

\begin{itemize}
\item \textbf{Y} the response 
\item \textbf{A}, \textbf{B}, \dots{} categorical covariates 
\item \textbf{G},\textbf{H}, \textbf{I}, \dots{} grouping factors for random effects 
\item \textbf{U}, \textbf{V}, \dots{} (skipping \textbf{Y}) continuous covariates
\end{itemize}

The timing results are saved in \href{http://json.org}{JSON (JavaScript
Object Notation)} files in the directory accessible as

\begin{Schunk}
\begin{Sinput}
> system.file("JSON",package="Timings")
\end{Sinput}
\end{Schunk}

within \pkg{R}. The directory name will end with
\texttt{./Timings/inst/JSON/} in the package source directory, for
example the result of cloning the
\href{https://github.com/Stat990-033/Timings}{github repository}. There
is one \texttt{.json} file for each data set. Each such file contains
results on timings of one or more models.

The \pkg{Timings} package for \proglang{R} provides a \texttt{retime}
function that takes the name of one of these JSON files and, optionally,
the name of a file with the updated timings. Similarly there are some
source files for Julia retimings.

%    \begin{Verbatim}[commandchars=\\\{\}]
%{\color{incolor}In [{\color{incolor}1}]:} \PY{n}{include}\PY{p}{(}\PY{l+s}{\PYZdq{}}\PY{l+s}{../julia%%/retime.jl}\PY{l+s}{\PYZdq{}}\PY{p}{)}
%        \PY{n}{retime}\PY{p}{(}\PY{l+s}{\PYZdq{}}\PY{l+s}{../JSON/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p%}{,}\PY{l+s}{\PYZdq{}}\PY{l+s}{/tmp/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p}{)}\PY{p}{;}
%       \PY{n}{retime}\PY{p}{(}\PY{l+s}{\PYZdq{}}\PY{l+s}{../JSON/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p}%{,}\PY{l+s}{\PYZdq{}}\PY{l+s}{/tmp/Alfalfa.json}\PY{l+s}{\PYZdq{}}\PY{p}{)}\PY{p}{;}
%\end{Verbatim}

    \begin{Verbatim}[commandchars=\\\{\}]
dsname => "Alfalfa"
    \end{Verbatim}

    \begin{Verbatim}[commandchars=\\\{\}]

        `parse` has no method matching parse(::Int64)
    while loading In[1], in expression starting on line 2

         in retime at /home/bates/git/Timings/inst/julia/retime.jl:17

    \end{Verbatim}

    The timing was repeated so that compilation time is not included in the
results. This repetition is only needed once per session.

A careful examination of these results shows that the main differences
in the Julia timings (the R timings are merely reported, not evaluated)
are that the \texttt{LN\_BOBYQA} and \texttt{LD\_MMA} optimizers are
much faster in the second run. This is because much of the code needs to
be compiled the first time that a derivative-free optimizer and a
derivative-based optimizer are used.

The names of the optimizers used with \texttt{lmm} are those from the
\href{https://github.com/JuliaOpt/NLopt.jl}{NLopt package} for
\textbf{Julia}. Names that begin with \texttt{LD\_} are gradient-based
methods. Names that begin with \texttt{LN\_} are derivative-free
methods. There is one other derivative-free method, \texttt{LN\_PRAXIS},
available in the \texttt{NLopt} package but, for some reason, it can
hang on very simple problems like this. Frequently we omit it.

The optimizers used with \texttt{lmer} include the \texttt{Nelder\_Mead}
optimizer built into the \texttt{lme4} package, the \texttt{bobyqa}
optimizer from the
\href{http://cran.rstudio.com/web/packages/minqa/index.html}{minqa
package}, the derivative-free optimizers from the
\href{http://cran.rstudio.com/web/packages/nloptr/index.html}{nloptr
package} and several optimizers from the
\href{http://cran.rstudio.com/web/packages/optimx/index.html}{optimx
pacakge}.

The \texttt{optimx:bobyqa} optimizer is just a wrapper around
\texttt{bobyqa} (bounded optimization by quadratic approximation) from
the \texttt{minqa} package and should provide results similar to those
from the \texttt{bobyqa} optimizer. For some reason the number of
function evaluations is not reported for the version in \texttt{optimx}.

The optimizers from \texttt{nloptr} (i.e.~those whose names begin with
\texttt{NLOPT\_LN\_}) use the same underlying code as do the similarly
named optimizers in the \texttt{NLopt} package for \textbf{Julia}. The
number of iterations to convergence should be similar for the same
underlying code, although not nessarily exactly the same because the
evaluation of the objective in \textbf{R} and in \textbf{Julia} may
produce slightly different answers. Also the convergence criteria in the
\textbf{Julia} version are more strict than those in the \textbf{R}
version

Also shown are the value of the criterion (negative twice the
log-likelihood, lower is better) achieved, the elapsed time and the
number of function and gradient evaluations. The \texttt{nopt} value is
the number of parameters in the optimization problem. \texttt{mtype} is
the model type in the \textbf{Julia} code. There are special methods for
solving the penalized least squares (PLS) problem, and for evaluating
the objective and its gradient when there is only one grouping factor
for the random effects. The model type is called \texttt{PLSOne}.

The \texttt{Alfalfa} example is a particularly easy one and all of the
optimizerws converge to an objective value close to -10.81023 in less
than 0.6 seconds.
